Propositions as Sessions, Mechanically

### 2015-04-13 paper/formalisation.tex
Semi-finalised issues and guidelines.

Note that the rules for polymorphism, $\exists$ and $\forall$, were not
included in the final Coq development since GV (\S~\ref{sec:gv}) does not
support polymorphism making the rules unnecessary for the
translation. Further, as noted by \citeauthor{Lee:2012}, support for
polymorphism in type variables creates an explosion of the infrastructure
requiring additional of opening and substituting definitions and lemmata.

@@

The translation of GV terms to CP processes is given, following
\citeauthor{Wadler:2014}, using continuation-passing style. In this style a
well-typed term in GV is parameterised by a name representing the channel
which communicates the result of the translation (a CP process). Due to the
locally nameless representation, the translation of GV terms required a
procedure to infer the type of terms in order to specify the binding type for
the corresponding CP process. An example, translating choice, is shown in
Figure~\ref{fig:trans}, note that I am using \citeauthor{Wadler:2014}'s style
but using an LNCQ notation for clarity. In order to construct the cut and
output processes, the type of \tm{M} and \tm{N} must be computed. So, the
translation takes the environment, in which the term is typed, as an argument
to query the type of free variables. Unfortunately, this query returns a type
of \coqe!option typ!  (\coqe!option! is equivalent to \tm{Maybe} in Haskell)
since it may not be found, but in fact since only well-typed GV terms are
translated it is guaranteed to return something!  Therefore, the definition is
unnecessarily cluttered with impossible error cases.

@@

\textbf{Representation.} As discussed in the previous sections representation
choice is a central issue in formalising programming language
metatheory. Decisions on how to define, for instance, types, calculi and
translations impact on future definitions and properties dependent on
them. Considerations range from exploiting the full expressivity of the
theorem prover by using dependent types in the case of Coq. to how to encode
the properties of interest and being able to assess accurately the most
fruitful encoding upfront. avoiding features that would provide a more faitful
relationship between formal and informal systems is desirable if those
features would unnecessarily complicated the reasoning. it is a tradeoff.

@@

In order to proceed with proving the goal, it is necessary to show that
$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
application of an appropriate lemma, and there are two possible ways to prove
this using plain Coq. One way is to apply the lemma to one of \key{PER} and
\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
namely backward reasoning, does not apply since the goal has nothing to do
uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert! tactic one
can utilise backward reasoning. An assertion temporarily focuses a different
goal (the asserted goal) which may then be tackled using tactics in the
standard way. Unfortunately, this approach requires the explicit writing of
the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script of the
lemma cluttering the proof script sensitive to change because \coqe!$\GD$P0!
and \coqe!$\GD$Q0! are emphemeral names introduced by prover.

@@

\textbf{Linear environments.} Linearity create issues with typing some
examples; the splitting of the environment was non-deterministic so required
side-conditions to enforce a certain splitting of binders. For example the
rule for associativity of cut. \todo{how to handle? is there a better way?}

\textbf{Effective Automation.} It is worth considering situations where
automation would simplify proofs. Automation is already an important
consideration in the interactive theorem proving community. Base rules and
rules with binders required permutation assumptions to allow any ordering of
the constituent contexts. a lot of the proofs for well-typed reductions
involved small amounts of permuting environments which were largely similar
across the proofs. while some simple lemmata and tactics where created to
handle the most tedious elements these did not make effective use of
hypotheses from the context. Improved proof search for permutations is
required to simplify such developments.

>>>>

Note that the rules for polymorphism, $\exists$ and $\forall$, were not
included in the final Coq development since GV (\S~\ref{sec:gv}) does not
support polymorphism making the rules unnecessary for the
translation. Further, as noted by \citeauthor{Lee:2012}, support for
polymorphism in type variables creates an explosion of the infrastructure
requiring additional open and substitution definitions and related lemmata.

@@

The translation of GV terms to CP processes is given, following
\citeauthor{Wadler:2014}, using continuation-passing style. In this style a
well-typed term in GV is parameterised by a name representing the channel
which communicates the result of the translation (a CP process). Due to the
locally nameless representation, the translation of GV terms required a
procedure to infer the type of terms in order to specify the binding type for
the corresponding CP process. An example, translating choice, is shown in
Figure~\ref{fig:trans}, unfortunately this definition is wrong because the
process P is not being opened and therefore during the proof we are unable to
reason about the typability of P opened with a sufficiently fresh variable the
reason is that P is a subprocess and therefore is not locally closed note that
I am using \citeauthor{Wadler:2014}'s style but using an LNCQ notation for
clarity. In order to construct the cut and output processes, the type of
\tm{M} and \tm{N} must be computed. So, the translation takes the environment,
in which the term is typed, as an argument to query the type of free
variables. Unfortunately, this query returns a type of \coqe!option typ!
(\coqe!option! is equivalent to \tm{Maybe} in Haskell) since it may not be
found, but in fact since only well-typed GV terms are translated it is
guaranteed to return something!  Therefore, the definition is unnecessarily
cluttered with impossible error cases.

@@

\textbf{Representation.} As discussed in the previous sections representation
choice is a central issue when formalising programming language
metatheory. Decisions on how to define, for instance, types, calculi and
translations impact on future definitions and properties dependent on
them. Considerations range from how much expressive power provided by the
theorem prover to exploit to the choice of encoding to use for the properties
of interest. Some of these decisions may impact the correspondence between the
formal and informal systems e.g. the empty choice rule and omission of
contraction in CP, and the omission of n-ary branch and choice from GV, and
such things are necessarily a tradeoff between simplicity and
adequacy.

Interplay between different decisions is an additional factor to consider. For
example, the representation choice for linearity makes environment splitting
non-deterministic. Applying inversion on well-typed process terms requires, in
some cases such as \coqe!assoc_1!, to enforce a certain split between
subprocesses. The choice of LNCQ representation forces a local closure
assumption in these cases. Alternative representations for the CP calculus,
such as separating the rules (link, cut, etc.) from the derivations
(associativity, reductions, etc.)~\cite{Tews:2013,Dawson:2010,Dawson:2002},
may ameliorate the handling of linearity by allowing one to explicate the
particular splitting for a derivation.

@@

In order to proceed with proving the goal, it is necessary to show that
$\GD$P0++$\GD$Q0 is unique. The proof follows from \key{PER} and \key{UN} upon
application of an appropriate lemma, and there are two possible ways to prove
this using plain Coq. One way is to apply the lemma to one of \key{PER} or
\key{UN} thereby replacing the chosen hypothesis with the conclusion of the
lemma (termed \textit{forward reasoning}). The opposite of forward reasoning,
namely backward reasoning, does not apply since the goal has nothing to do
uniqueness of $\GD$P0++$\GD$Q0. However, using the \coqe!assert! tactic one
can utilise backward reasoning. An assertion temporarily focuses a different
goal (the asserted goal) which may then be tackled using tactics in the
standard way. Unfortunately, this approach requires the explicit writing of
the goal, namely \coqe!uniq ($\GD$P0++$\GD$Q0)!, into the proof script of the
lemma cluttering the proof script sensitive to change because \coqe!$\GD$P0!
and \coqe!$\GD$Q0! are ephemeral names introduced by prover.

@@

\textbf{Effective Automation.} It is worth considering situations where
automation would simplify proofs. Automation is already an important
consideration in the interactive theorem proving community. Examples of
automation from the current development have been highlighted, but particular
emphasis should be placed on tactics involving non-trivial proof search of the
context because these tend to be more widely applicable.
